---
title: "Mini Data Analysis Milestone 2"
output: html_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(ggplot2)
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

```{r}
# Summary of *apt_buildings* data set for reference:

str(apt_buildings)
```

#### 1.1.1. Research question 1: Investigating the relationship between building age and building height

Investigate the relationship between building age and building height. One could hypothesize that newer buildings might tend to be taller as construction technology has improved and urban density has increased. 


#### 1.1.2. Research question 2: Investigating the relationship between building height, number of units, and balconies

Investigate the relationship between building height, number of units, and balconies. One could hypothesize that, given constant area per floor and area per unit, building height and number of units would display a positive correlation. Similarly, one could hypothesize that higher buildings or buildings with more units are more likely to have balconies. 


#### 1.1.3. Research question 3: Investigating building age dependency of air conditioning type

Investigate the relationship between building age and air conditioning type. For example, one could hypothesize that newer buildings might more frequently have central air conditioning, while older buildings might more frequently lack air conditioning. 


#### 1.1.4. Research question 4: Investigating the relationship between building height and ward

Investigate the relationship between building height and city ward. One could hypothesize that buildings in denser downtown wards might tend to be higher, while buildings in outlying suburban wards might tend to be shorter. 


### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

#### 1.2.1. Research question 1: Investigating the relationship between building age and building height

This question will be investigated using options 3 and 5. 

For option 3 (creating a categorical variable by grouping a numerical variable), building age will be grouped into categories by 25-year period. Groups will be: Pre-1900, 1900-1925, 1925-1950, 1950-1975, 1975-2000, and Post-2000.

For option 5 (creating a graph of the summarized variables with 2+ geom layers), a box plot will be created for building height in each age category. This plot will be overlaied with the raw datapoints.


#### 1.2.2. Research question 2: Investigating the relationship between building height, number of units, and balconies

This question will be investigated using options 1 and 7. 

For option 1 (compute the range, mean, and two other summary stats of one categorical variable), the summary stats will be calculated for the categories of balcony and no balcony. 

For option 7 (graph with customized alpha transparency), a scatter plot will be created with number of units on the x axis and building height on the y axis. Two seperate graphs will be created for buildings with balconies and no balconies. Alpha transparency will be applied to the datapoints to demonstrate density of observations. 


#### 1.2.3. Research question 3: Investigating building age dependency of air conditioning type

This question will be investigated using options 1 and 5. 

For option 1 (compute the range, mean, and two other summary stats of one categorical variable), the summary stats will be calculated for the categories of air conditioning type.

For option 5 (creating a graph of the summarized variables with 2+ geom layers), a box plot will be created for building height in each age category. This plot will be overlaied with the raw datapoints.


#### 1.2.4. Research question 4: Investigating the relationship between building height and ward

This question will be investigated using options 2 and 6.

For option 2 (calculate number of observations for one categorical variable), the number of buildings in each ward will be computed. 

For option 6 (chosen graph with one log axis and "pretty labels), the distribution of buildings in each ward will be shown using a ridge plot. A horizontal log axis will be used to better display the higher density of observations at low numbers of floors. 


<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->


<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->


<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 